---
title: "Turner & O’Halloran et al., *Nature*, 2021"
output: html_notebook
---

Bulk BCR-seq

Author: Julian Q. Zhou

Docker container: `julianqz/wu_pub:r_4.1.0`

Visualizations are outputted as pdfs.


# Load packages & helper functions

```{r}
# modify the next line to set working directory accordingly
# setwd("[insert_your_working_directory_here]")

# get_bcr_network
source("./c2b2/bcr_graphs.R")

# scale_values, mystar
source("./c2b2/misc_ops.R")

# vioplotLst
source("./c2b2/vis_vioplot.R")
```

```{r}
library(igraph)
library(vioplot)
```

```{r}
versions::installed.versions(c("igraph", "vioplot"))
```


# Setup

```{r}
# filenames
fn_db = "WU368_turner_et_al_nature_2021_bcr.tsv.gz"
fn_nat = "mufreq_turner_et_al_nature_2020.tsv.gz"

# column names
col_clone = "clone_id"
col_mab_clone = "mab_clone"
col_seq_id = "sequence_id"
col_seq_type = "seq_type"
col_cdr3 = "cdr3"
col_mf = "nuc_RS_freq_19_312"
col_size = "duplicate_count"
col_donor = "donor"

# Hamming distance threshold for computing edges in networks
threshold = 0.15

# participant IDs
DONORS = paste0("368-", c("07", "20", "22"))

# sequence IDs of heavy chains of cross-reactive S-binding mAbs
mab_cross = c("368-22_m3_2021-05-22@368.22.28.GC.1B08.HC", 
              "368-22_m3_2021-05-22@368.22.28.GC.1G10.HC", 
              "368-22_m3_2021-05-22@368.22.28.GC.1C02.HC")
```


# Load BCR Data

```{r}
# processed BCR data from Turner & O’Halloran et al., Nature, 2021
db_zz = gzfile(fn_db,'rt')
db = read.table(db_zz, header=T, sep="\t", stringsAsFactors=F)
close(db_zz)

# db contains B cell clones from all donors
# The same clone ID could exist in two different donors
# But those are two separate clones
# Make that distinction by appending donor ID in front of $clone_id
db[[col_clone]] = paste0(db[[col_donor]], "@", db[[col_clone]])

# clone IDs of clones containing S-binding mAbs
clones_mab = unique(db[[col_clone]][db[[col_seq_type]]=="mab"])

# TRUE if a sequence is in a mAb clone
db[[col_mab_clone]] = db[[col_clone]] %in% clones_mab
```


```{r}
# Recalculate SHM frequency to avoid numerical problems affecting $nuc_RS_freq_19_312
# that arise from writing and reading tsv's

# E.g. a "0.013108614" could become "0.013108615" after being exported and read back in

# Doing so ensures that the exact p-values can be reproduced
db[[col_mf]] = db[["nuc_RS_19_312"]] / db[["nuc_denom_19_312"]]
```


```{r}
# SHM frequency data from Turner & Zhou & Han et al., Nature, 2020
db_nat_zz = gzfile(paste0("./assets/", fn_nat),'rt')
db_nat = read.table(db_nat_zz, header=T, sep="\t", stringsAsFactors=F)
close(db_nat_zz)

# extract SHM frequency of the naive compartment
nat_naive = db_nat[["MF"]][db_nat[["SUBSET"]]=="Naive"]
```


# Visualization of B cell clones 

* Fig 3b
* Extended Data Fig 3b

### Fig 3b (Participants 07, 22) & Extended Data Fig 3b (Participant 20)

```{r}
### prep

# add star to shapes
# set border to transparent
add_shape("star", clip=shape_noclip, plot=mystar, 
          parameters=list(vertex.norays=5, vertex.frame.color="transparent"))


# global range of col_size

lst_size = vector(mode="list", length=length(DONORS))
names(lst_size) = DONORS

for (d in DONORS) {
    # subset to mAb clones
    db_mab = db[db[[col_donor]]==d & db[[col_mab_clone]], ]
    stopifnot(nrow(db_mab)>0)
    
    lst_size[[d]] = db_mab[[col_size]]
}

# [1, 40]
(global_range_size = range(unlist(lst_size), na.rm=T))


### visualize

# For the published version, mAb IDs were not printed by code; instead, they 
# were added next to each clone later in Adobe Illustrator.
# However, it would help to know the mAb to which each clone corresponds
# Toggle ``print_mab_ids to generate plots with or without mAb IDs printed

print_mab_ids = FALSE
#print_mab_ids = TRUE

for (d in DONORS) {
    
    cat("\n", d, "\n")
    
    # subset to mAb clones
    db_mab = db[db[[col_donor]]==d & db[[col_mab_clone]], ]
    stopifnot(nrow(db_mab)>0)
    
    # simplify sequence ID of mAb heavy chains
    # E.g.:
    # before: "368-07_m1_2021-05-22@368.07.28.GC.2C08.HC"
    # after: 2C08
    mab_ids = sapply(db[[col_seq_id]][db[[col_seq_type]]=="mab" & db[[col_donor]]==d],
                     function(s){ strsplit(s, "\\.")[[1]][5] }, USE.NAMES=F)
    
    ### generate network
    
    g = get_bcr_network(db=db_mab, col_clone=col_clone, col_seq=col_cdr3, 
                        threshold=threshold, nt_or_aa="nt", 
                        bool_norm=T, bool_mst=T)
    
    ### configure network aesthetics
    
    ## vertices
    
    # label mAbs?
    if (print_mab_ids) {
        V(g)$label[db_mab[[col_seq_type]]=="mab"] = mab_ids
    } else {
        V(g)$label = rep("", length(V(g)))
    }
    
    ## shape
    # ?igraph::shapes
    
    # by isotype
    shape_uniq = c("IGHG"="circle", "IGHA"="star", "IGHM"="square")
    stopifnot(all(db_mab[["isotype"]] %in% names(shape_uniq)))
    
    vec_shapes = shape_uniq[db_mab[["isotype"]]]
    V(g)$shape = vec_shapes
    
    ## color
    
    col_color = "sorting"
    color_uniq = c("GC"="blue", "PB"="red", "IgDlo"="darkgoldenrod2")
    
    vec_color = color_uniq[db_mab[[col_color]]]
    names(vec_color) = NULL
    V(g)$color = vec_color
    
    # special shade of blue for mAbs (all mAbs are GC per experimental design)
    V(g)$color[db_mab[[col_seq_type]]=="mab"] = "cyan"
    
    ## border
    V(g)$frame.color = "transparent"
    
    ## size
    
    # there could be NA in col_size
    # convert any NA to 1
    vec_size = ifelse(is.na(db_mab[[col_size]]), 1, db_mab[[col_size]])
    print(summary(vec_size))
    
    # convert to be on a scale that is common amongst all donors
    # with keep_NA=F, any NA would be set to min_scale
    vec_scale = scale_values(vec_orig=vec_size, 
                             max_orig=global_range_size[2],
                             min_orig=global_range_size[1],
                             max_scale=12.5, min_scale=4.5, keep_NA=FALSE)
    print(summary(vec_scale))
    
    V(g)$size = vec_scale
    
    ## edges
    
    E(g)$color = "gray85"
    
    ## layout
    
    # defaults to layout_nicely
    coords=NULL
    
    ### plot
    
    fn_p = paste0("network_", d, ".pdf")
    pdf(fn_p, width=5, height=5)
    par(mar=c(0.2,0.2,0.2,0.2)) # BLTR
    
    set.seed(8475240)
    plot(g, layout=coords)
    box()
    
    dev.off()
    
    # clean up
    rm(g, db_mab, mab_ids, vec_scale)
}

```


# Visualization of SHM frequency

* Fig 3c
* Fig 3d

*The code-generated plots from below were prototypes based on which the published 
versions were generated using GraphPad Prism v8.*

### Fig 3c (Naive vs. S-binding Clones)

```{r}
### prep

# get SHM frequency of mAb clones in each donor

# list of data.frames
# each data.frame corresponds to a donor
# (this format helps with plotting cross vs. non-cross reactive later)
lst_df = vector(mode="list", length=length(DONORS))
names(lst_df) = DONORS

for (d in DONORS) {
    # clone IDs of mAb clones from current donor
    clones_mab = unique(db[[col_clone]][db[[col_seq_type]]=="mab" & db[[col_donor]]==d])
    stopifnot(length(clones_mab)>0)
    
    cur_db = db[db[[col_clone]] %in% clones_mab,
                c(col_donor, col_clone, col_mf, col_seq_id)]
    
    lst_df[[d]] = cur_db
}

# list of vectors
# each vector contains SHM frequencies from a different donor
lst_mf = lapply(lst_df, function(x){ x[[col_mf]] })

# config plot
global_mf_range = range(c(unlist(lst_mf), nat_naive), na.rm=T)

lst_all = c(list("Naive"=nat_naive), lst_mf)


### visualize

fn_p = "fig_3c.pdf"

pdf(fn_p, width=5, height=5)
par(mar=c(2.7,4.5,0.2,0.2)) # BLTR

set.seed(9375243)
vioplotLst(vecLst=lst_all, vioYmax=global_mf_range[2]*1.05, vioYmin=0, 
           xLas=1, yLas=2, 
           xNames=names(lst_all), xTick=TRUE, xlab="", 
           ylab="Nucleotide mutation frequency in IGHV", title="", 
           countCex=0.7, xaxisCex=0.85, yaxisCex=1, mainCex=1.2,
           yHoriz=NULL, yHorizCol="firebrick2",
           vioBkgCol=NA, vioMedCol="hotpink", vioRecCol="darkgoldenrod2",
           vioMedPch=18, vioMedStyle="point", vioBoxWidth=0.2, vioWhiskerCol="black",
           useBoxplot=T, showPoints=T, 
           pointsPch=rep(16, length(lst_all)), 
           pointsCol=c("gray70", "forestgreen", "purple", "purple", "purple"), 
           pointsCex=0.3, 
           pointsJitterAmount=0.4, pointsJitterBool=TRUE)

dev.off()

# "Kruskal–Wallis test with Dunn’s post-test between naive B cells and 
# S-binding clones" performed in GraphPad Prism v8
```


### Fig 3d (Cross-reactive vs. Non-cross-reactive S-binding clones)

```{r}
### prep

df_mf = do.call(rbind, lst_df)

# verify that all cross-reactive S-binding mAbs are in db
stopifnot(all(mab_cross %in% df_mf[[col_seq_id]]))

# clones containing cross-reactive mAbs
mab_cross_clones = unique(df_mf[[col_clone]][match(mab_cross, df_mf[[col_seq_id]])])

# TRUE if a sequence is in one of the clones containing cross-reactive mAbs
bool_cross_clones = df_mf[[col_clone]] %in% mab_cross_clones

lst_mf_2 = list("Cross-reactive"=df_mf[[col_mf]][bool_cross_clones],
                "Non-cross-reactive"=df_mf[[col_mf]][!bool_cross_clones])


### visualize

fn_p = "fig_3d.pdf"
pdf(fn_p, width=5, height=5)
par(mar=c(2.7,4.5,0.2,0.2)) # BLTR

set.seed(9375243)
vioplotLst(vecLst=lst_mf_2, vioYmax=global_mf_range[2]*1.05, vioYmin=0, 
           xLas=1, yLas=2, 
           xNames=names(lst_mf_2), xTick=TRUE, xlab="", 
           ylab="Nucleotide mutation frequency in IGHV", title="", 
           countCex=0.7, xaxisCex=0.85, yaxisCex=1, mainCex=1.2,
           yHoriz=NULL, yHorizCol="firebrick2",
           vioBkgCol=NA, vioMedCol="hotpink", vioRecCol="darkgoldenrod2",
           vioMedPch=18, vioMedStyle="point", vioBoxWidth=0.2, vioWhiskerCol="black",
           useBoxplot=T, showPoints=T, 
           pointsPch=rep(16, length(lst_mf_2)), 
           pointsCol=c("gray70", "gray70"), #* 
           pointsCex=0.3, 
           pointsJitterAmount=0.4, pointsJitterBool=TRUE)

dev.off()

# Mann-Whitney
# 9.410932e-30
wilcox.test(x=lst_mf_2[["Cross-reactive"]], y=lst_mf_2[["Non-cross-reactive"]], 
            alternative="two.sided", paired=F)$p.value
```

