---
title: "Malladi & Jaiswal et al., *Science Immunology*, 2025 (repertoire)"
output: html_notebook
---

Author: Julian Q. Zhou

Docker container: `julianqz/wu_pub:r_4.1.0`

Visualizations are outputted as pdfs or pngs.

# Load packages & helper functions

```{r}
# modify the next line to set working directory accordingly
# setwd("[insert_your_working_directory_here]")

# summarize_clone
source("./c2b2/bcr_clone_info.R")

# tabulate_vj
source("./c2b2/bcr_vj.R")

# circos_clonal_overlap
source("./c2b2/bcr_clonal_overlap_circos_v2.R")

# prep_for_alluv, prep_for_paired_mf, get_stat_paired_mf
source("./c2b2/bcr_alluvial.R")

# vioplotLst
source("./c2b2/vis_vioplot.R")
```

```{r}
library(tibble)
library(ggplot2)
```

```{r}
versions::installed.versions(c("ggplot2", "circlize", "vioplot"))
```


# Setup

```{r}
### filenames
fn_db = "WU382_malladi_et_al_sci_imm_2025_bcr_heavy.tsv.gz"
fn_umap = "WU382_malladi_et_al_sci_imm_2025_gex_b_cell_umap.tsv.gz"
fn_binding = "WU382_malladi_et_al_sci_imm_2025_mab_binding.tsv"

### column names

col_iso = "isotype"
col_comp = "compartment"
col_tp = "timepoint"
# [compartment]_[timepoint]
col_comp_tp = "compartment_timepoint"
col_expressed_id = "expressed_id"

# B cell clonal lineage ID
col_clone_id = "clone_id"
# TRUE if a clonal member of a S+ heavy chain-based B cell clone
col_s_pos_clone = "s_pos_clone"


# transcriptomics-based cluster annotation
col_anno = "anno_leiden_0.25"

# replacement and silent mutation frequency
col_mf = "nuc_RS_freq_1_312"

col_tissue = "tissue"
col_seq_id = "sequence_id"
col_donor = "donor"
col_sample = "sample"

### participants
DONORS = paste0("382-", c("65","67","69","70","71"))
```


# Load BCR data

```{r}
# db_all contains heavy chains from all clones regardless of S specificity
db_all_zz = gzfile(fn_db_bd,'rt')
db_all = read.table(db_all_zz, header=T, sep="\t", stringsAsFactors=F)
close(db_all_zz)

# Subset to S+ clones
db = db_all[db_all[[col_s_pos_clone]], ]
```

```{r}
# Recalculate SHM frequency to avoid numerical problems affecting $nuc_RS_freq_19_312
# that arise from writing and reading tsv's

# E.g. a "0.013108614" could become "0.013108615" after being exported and read back in

# Doing so ensures that the exact p-values reported in the paper can be reproduced
db[[col_mf]] = db[["nuc_RS_1_312"]] / db[["nuc_denom_1_312"]]
```


```{r}
# Create a [compartment]_[timepoint] column
db[[col_comp_tp]] = paste0(db[[col_comp]], "_", db[[col_tp]])
```


# Clone-level summary

First, we create a `data.frame` in which each row represents a clone, and columns
tabulate the number of sequences in a compartment and in a compartment-timepoint
combination.

```{r}
clone_info = summarize_clone(db=db, col_clone=col_clone_id, 
                             col_vec=c(col_comp, col_comp_tp),
                             order_by_size="none")

# Extract participant ID from clone ID, which has the format [donor]@[clone]
clone_info[[col_donor]] = sapply(clone_info[[col_clone_id]],
                                 function(s){strsplit(s, "@")[[1]][1]})
```


# Circos plots for clonal overlap

* Fig S3

`circos_clonal_overlap-v2` produces a circos plot, and prints out (i) the number and
percentage of clones with clonal overlap, and (ii) the number of sequences.

Starting with the arc at the 12 o'clock position and going clockwise, the identities of 
the arcs correspond to the rows in the printed table from top to bottom.

### How to read the circos plots

* An arc represents BCR sequences from a B cell compartment or 
compartment-timepoint combination
* The length of an arc corresponds to the number of BCR sequences 
represented by the arc
* Chords represent clonal connections spanning at least two arcs
* The widths of chord ends correspond to the number of sequences from the respective arc
* Chords are colored based on clonal overlap

### Visualization

```{r}
# Fig S3 - clonal overlap across lymph node and peripheral blood compartments

# arcs: compartment-timepoint combinations
vec_sectors_1 = c("MBC_PBMC_d0", "PB_d8", "GC_d57", "MBC_PBMC_d121")

# color arcs according to compartment
vec_sectors_color_uniq_1 = c("PB"="firebrick2",
                             "GC"="dodgerblue2",
                             "LNPC"="forestgreen",
                             "MBC_PBMC"="plum2")
vec_sectors_type_1 = c("MBC_PBMC","PB","GC","MBC_PBMC")
vec_sectors_color_1 = vec_sectors_color_uniq_1[vec_sectors_type_1]

# color chords according to clonal overlap
vec_color_overlap = c("GC@PB"="purple",
                      "GC@MBC_PBMC"="forestgreen")
color_rest = "gray40"

# width of gaps between arcs
vec_gaps_1 = rep(4, length(vec_sectors_1))

# column in sequence data.frame that contains arc labels
col_sector_1 = col_comp_tp

fn_sink = "fig_S3_info.txt"
sink(fn_sink)

for (d in DONORS) {
    cat("\n", d, "\n")
    # subset to current donor
    d_db = db[db[[col_donor]]==d, ]
    d_clone_info = clone_info[clone_info[[col_clone_id]] %in% d_db[[col_clone_id]], ]
    
    d_fn = paste0("fig_S3_", d, ".pdf")
    pdf(d_fn, width=5, height=5)
    
    # vis
    circos_clonal_overlap_v2(vec_sectors=vec_sectors_1, 
                             vec_sectors_type=vec_sectors_type_1, 
                             vec_sectors_color=vec_sectors_color_1, 
                             vec_gaps=vec_gaps_1,  
                             col_sector=col_sector_1, 
                             col_clone_id=col_clone_id, 
                             df_seq_data=d_db, 
                             df_clone_info=d_clone_info, 
                             color_overlap=vec_color_overlap,
                             color_rest=color_rest,
                             color_mono=FALSE)
    
    dev.off()
}

sink()
```

# V gene usage heatmap (Fig S2)

```{r}
# tabulate IGHV usage at the clone-level
idx_cl_h = match(clone_info[[col_clone_id]], db[[col_clone_id]])
stopifnot(!any(is.na(idx_cl_h)))
stopifnot(all.equal(clone_info[[col_clone_id]], db[[col_clone_id]][idx_cl_h]))
clone_info[["h_germline_v_call"]] = db[["germline_v_call"]][idx_cl_h]
clone_info[["h_germline_j_call"]] = db[["germline_j_call"]][idx_cl_h]

lst_vj_h = tabulate_vj(vec_v=clone_info[["h_germline_v_call"]], 
                       vec_j=clone_info[["h_germline_j_call"]], 
                       vec_subj=clone_info[[col_donor]],
                       anno_separator=",")

# x-axis: IGHV
# y-axis: participant
# heatmap: num of clones

lst_v_common_h = vector(mode="list", length=length(DONORS))
names(lst_v_common_h) = DONORS

for (cur_d in DONORS) {
    cur_tab = lst_vj_h[["common"]][[cur_d]]
    cur_vec_uniq_v = unique(cur_tab[["v"]])
    
    cur_vec_cols_df = c("subj", "v", "count")
    cur_df = data.frame(matrix(NA, nrow=length(cur_vec_uniq_v),
                               ncol=length(cur_vec_cols_df)))
    colnames(cur_df) = cur_vec_cols_df
    
    cur_df[["subj"]] = cur_d
    for (i_row in 1:nrow(cur_df)) {
        cur_v = cur_vec_uniq_v[i_row]
        cur_df[["v"]][i_row] = cur_v
        cur_df[["count"]][i_row] = sum( cur_tab[["count"]][cur_tab[["v"]]==cur_v] )
    }
    stopifnot( sum(cur_df[["count"]]) == sum(cur_tab[["count"]]) )
    
    lst_v_common_h[[cur_d]] = cur_df
}

df_v_common_h = do.call(rbind, lst_v_common_h)
rownames(df_v_common_h) = NULL

stopifnot(sum(df_v_common_h[["count"]])==nrow(clone_info))

# vis

# https://colorbrewer2.org/#type=sequential&scheme=GnBu&n=9
palette_num_donor = c('#f7fcf0','#e0f3db','#ccebc5','#a8ddb5','#7bccc4','#4eb3d3','#2b8cbe','#0868ac','#084081')


g = ggplot(df_v_common_h, 
                              aes(x=v, y=subj, fill=count)) +
    # white border with thickness
    geom_tile(color="white", size=0.25) +
    # remove x and y axis labels
    labs(x="", y="")+
    # remove extra space
    scale_y_discrete(expand=c(0,0)) +
    scale_x_discrete(expand=c(0,0)) +
    # color scale
    scale_fill_gradient(low=palette_num_donor[1], high=palette_num_donor[9]) +
    # set a base size for all fonts
    theme_grey(base_size=8) +
    # theme
    theme(
        #legend.position=c(1.02,0.1), 
        # remove legend title
        legend.title=element_blank(),
        # legend size
        legend.key.size=unit(0.25, "cm"),
        # remove plot background
        plot.background=element_blank(),
        # remove plot border
        panel.border=element_blank(),
        # remove ticks
        axis.ticks=element_blank(),
        # rotate x axis label
        axis.text.x=element_text(angle=90, hjust=1, vjust=0.5),
        # margin
        plot.margin=margin(0.05,0,0,0,"cm") # TRBL
)

fn = "fig_S2F.pdf"
ggsave(g, filename=fn,
       height=1.5, width=6, units="in")
rm(g)
```

# Isotype

```{r}
vec_uniq_iso = paste0("IGH", c("D","M","A","G"))
vec_uniq_iso_color = c("IGHM"="darkgoldenrod2", "IGHD"="black",
                       "IGHA"="forestgreen", "IGHG"="hotpink")
vec_uniq_iso_lab = gsub(pattern="^IGH", replacement="Ig", x=vec_uniq_iso)

vec_tab_iso_gcb = table(db[[col_iso]][db[[col_comp_tp]]=="GC_d57"],
                        useNA="ifany")[vec_uniq_iso]

vec_tab_iso_gcb_perc = vec_tab_iso_gcb/sum(vec_tab_iso_gcb)*100
iso_perc_max_plot = 100 

fn = "fig_x_A.pdf"
pdf(fn, width=4, height=4)
par(mar=c(2,4.5,2,1)) # BLTR
barplot(vec_tab_iso_gcb_perc,
        ylim=c(0,iso_perc_max_plot),
        col=vec_uniq_iso_color[vec_uniq_iso], border=NA,
        names.arg=vec_uniq_iso_lab, ylab="% of d57 GC B cells",
        main="GC B cell")
dev.off()

```


# SHM frequency

```{r}
fn = "fig_x_B.pdf"
pdf(fn, width=5, height=5)
par(mar=c(2,4.5,1,1)) # BLTR
boxplot(db[[col_mf]][db[[col_comp_tp]]=="PB_d8"],
        db[[col_mf]][db[[col_comp_tp]]=="GC_d57"],
        db[[col_mf]][db[[col_comp_tp]]=="MBC_PBMC_d121"],
        col=c("firebrick2","dodgerblue2","plum2"),
        ylab="IGHV nucleotide mutation frequency",
        axes=F)
axis(2)
axis(1, tick=F,
     at=1:3, labels=c("\nd8\nPB", "d57\nGCB",  "d121\nMBC"))
box()
dev.off()

# p-values
# 3.4382e-24
wilcox.test(db[[col_mf]][db[[col_comp_tp]]=="PB_d8"],
            db[[col_mf]][db[[col_comp_tp]]=="GC_d57"])$p.value
# 1.476817e-56
wilcox.test(db[[col_mf]][db[[col_comp_tp]]=="PB_d8"],
            db[[col_mf]][db[[col_comp_tp]]=="MBC_PBMC_d121"])$p.value
# 1.404774e-14
wilcox.test(db[[col_mf]][db[[col_comp_tp]]=="GC_d57"],
            db[[col_mf]][db[[col_comp_tp]]=="MBC_PBMC_d121"])$p.value
```


# Clone-wise SHM frequency comparisons between compartments

* Fig ? (PB vs. GC B cells)

### How to read the clone-wise SHM frequency plots

* Each line corresponds to a B cell clone
* Each end of the line corresponds to the median SHM frequency of the clonal 
members in a given compartment or compartment-timepoint combination

### Preparation

```{r}
# all compartment-timepoint combinations to consider
alluv_cols = c("PB_d8", "GC_d57")

# regex to extract what labels define overlap
vec_overlap_regex = c("PB_d8", "GC_d57")

# `prep_for_alluv` returns `alluv_df` and `alluv_lst_cl_id` (explained below)
alluv_prep = prep_for_alluv(alluv_cols=alluv_cols, 
                            df_clone_info=clone_info,
                            col_df_clone_id=col_clone_id,
                            vec_overlap_regex=vec_overlap_regex)

# `alluv_df` 
# (i) exhaustively lists all binary combinations of all compartment-timepoint's
# (ii) counts the number of clones belonging to each binary combination
# (iii) determines whether each binary combination constitutes an overlap

# E.g.:
#   PB_d8 GC_d57 count overlap
# 1  TRUE   TRUE    50    TRUE
# 2 FALSE   TRUE   299   FALSE
# 3  TRUE  FALSE     6   FALSE
# 4 FALSE  FALSE   109   FALSE
alluv_df = alluv_prep[["alluv_df"]]

# `alluv_lst_cl_id`
# For each binary combination (row) in `alluv_df`,
# record the clone IDs of the clones belonging to that binary combination
alluv_lst_cl_id = alluv_prep[["alluv_lst_cl_id"]]

# compartment-timepoint's to visualize
lst_pairs = list(c("PB_d8", "GC_d57"))

# `prep_for_paired_mf` returns `vec_facets`, `pairs_df_wide/long` (explained below)
prep_pairs = prep_for_paired_mf(lst_pairs=lst_pairs, 
                                alluv_cols=alluv_cols, 
                                alluv_df=alluv_df, 
                                alluv_lst_cl_id=alluv_lst_cl_id,
                                df_data=db, 
                                col_df_clone_id=col_clone_id, 
                                col_df_mf=col_mf, 
                                col_df_label=col_comp_tp)

# A wide-format data.frame
# Each row corresponds to a clone
# `a_ident` and `b_ident` specifies the compartment-timepoint's of interest
# `a_val` and `b_val` records the clonal median SHM frequency of `a_ident` and `b_ident` resp.

# E.g.:
#             facet        clone_id a_ident b_ident      a_val      b_val
# 1  PB_d8 + GC_d57  382-65@G1186_2   PB_d8  GC_d57 0.05614035 0.07719298
# 2  PB_d8 + GC_d57  382-65@G1205_1   PB_d8  GC_d57 0.05614035 0.08070175
pairs_df_wide = prep_pairs[["pairs_df_wide"]]

# The same data.frame in long-format; used by ggplot2
pairs_df_long = prep_pairs[["pairs_df_long"]]

# create an artificial dummy column to control line colour
pairs_df_long[["hack"]] = factor("dummy")
```


### Visualization

```{r}
# y-axis upper limit
ymax_gg = 0.2

# line color
col_line = scales::alpha("purple", 0.35)

g = ggplot(pairs_df_long, 
    aes(x=ident, y=val, group=clone_id)) + 
    geom_line(aes(color=hack)) +
    ylim(0, ymax_gg*1.02) +
    geom_point(aes(color=ident))+
    scale_color_manual(values=c(
        "PB_d8"="firebrick2", "GC_d57"="dodgerblue2", 
        "dummy"=col_line)) +
    labs(y="IGHV nucleotide mutation frequency") +
    theme_bw() +
    theme(axis.title.x=element_blank()) +
    theme(axis.text.x=element_text(size=13)) +
    theme(axis.text.y=element_text(size=13)) +
    theme(axis.title=element_text(size=13)) +
    theme(strip.background=element_rect(fill="transparent"),
          strip.text.x=element_text(size=13)
    ) +
    theme(legend.position="none") +
    theme(panel.grid.major=element_blank())

fn = "fig_x_C.pdf"
ggsave(filename=fn, plot=g, device="pdf", 
       width=3, height=4) 
rm(g)

# stats
# (i) print out, for each facet, 
#  - the number of clones
#  - the median SHM frequency of each compartment-timepoint
#    (a_val is always plotted on the left; b_val always on the right)
# (ii) perform two-sided, paired, Mann-Whitney tests
#  - correction for multiple testing by Benjamini & Hochberg
pairs_p = get_stat_paired_mf(vec_facets=prep_pairs[["vec_facets"]], 
                             pairs_df_wide=pairs_df_wide,
                             p_adj_method="BH")

pairs_p[["vec_p_unadj"]]
```


# Cell type

```{r}
df_binding = read.table(fn_binding, sep="\t", header=T, stringsAsFactors=F)

head(df_binding)

stopifnot(all(df_binding[["HC_ID"]] %in% db_all[[col_expressed_id]]))

### broadly binding
# 1 for all of {WA1_RBD BA1_RBD BQ11_RBD XBB15_RBD}
idx_bb = which(df_binding[["WA1_RBD"]]==1 &
               df_binding[["BA1_RBD"]]==1 & 
               df_binding[["BQ11_RBD"]]==1 &
               df_binding[["XBB15_RBD"]]==1)
# 106
length(idx_bb) 

### broadly neutralizing
# Neut for "Authentic_Virus_Neut" AND
# <1000 for IC50_WA1/BA1/XBB15
idx_bn = which(df_binding[["Authentic_Virus_Neut"]]=="Neut" &
               df_binding[["IC50_WA1_D614G"]]<1000 & 
               df_binding[["IC50_BA1"]]<1000 &
               df_binding[["IC50_XBB15"]]<1000)
# 10
length(idx_bn) 

### WA1-only binder
idx_wa1_only_binder = which(df_binding[["WA1_Spike"]]==1 &
                            df_binding[["BA1_Spike"]]==0)
# 85
length(idx_wa1_only_binder) 

### WA1-only neutralizer
idx_wa1_only_neut = which(df_binding[["Authentic_Virus_Neut"]]=="Neut" &
                          df_binding[["IC50_WA1_D614G"]]<1000 & 
                          df_binding[["IC50_BA1"]]>=1000 &
                          df_binding[["IC50_XBB15"]]>=1000)
# 5
length(idx_wa1_only_neut) 


### cell types, LN vs blood

vec_cell_type_uniq_ln = paste0(c("GC","LNPC","MBC_LN","Naive"), "_d57")
vec_cell_type_uniq_pbmc = c("MBC_PBMC_d0", "PB_d8", "MBC_PBMC_d121")

vec_cell_type_uniq_ln_color = c("dodgerblue2", "forestgreen", "plum2", "darkgoldenrod2")
vec_cell_type_uniq_pbmc_color = c("plum2", "firebrick2", "plum2")

vec_cell_type_uniq_ln_label = paste0("d57 ", c("GCB", "LNPC", "MBC", "Naive"))
vec_cell_type_uniq_pbmc_label = c("d0 MBC", "d8 PB", "d121 MBC")


for (cur_type in c("bb", "bn", "wa1_b", "wa1_n")) {
    
    cat(cur_type, "\n")
    
    # wrt df_binding
    if (cur_type=="bb") {
        cur_idx_1 = idx_bb
        cur_title = "Broadly binding"
    } else if (cur_type=="bn") {
        cur_idx_1 = idx_bn
        cur_title = "Broadly neutralizing"
    } else if (cur_type=="wa1_b") {
        cur_idx_1 = idx_wa1_only_binder
        cur_title = "WA1-binding only"
    } else if (cur_type=="wa1_n") {
        cur_idx_1 = idx_wa1_only_neut
        cur_title = "WA1-neutraling only"
    }
    
    cur_vec_mabs = df_binding[["HC_ID"]][cur_idx_1]

    stopifnot(all(cur_vec_mabs %in% db_all[[col_expressed_id]]))
    stopifnot(all(cur_vec_mabs %in% db[[col_expressed_id]]))
    
    # wrt db
    cur_idx_2 = match(cur_vec_mabs, db[[col_expressed_id]])
    stopifnot(!any(is.na(cur_idx_2)))
    stopifnot(all.equal(cur_vec_mabs, db[[col_expressed_id]][cur_idx_2]))
    
    cur_vec_mabs_clone_id = db[[col_clone_id]][cur_idx_2]
    
    cur_db_ln = db[db[[col_clone_id]] %in% cur_vec_mabs_clone_id &
                   db[[col_tissue]]=="LN", ]
    cur_db_pbmc = db[db[[col_clone_id]] %in% cur_vec_mabs_clone_id &
                     db[[col_tissue]]=="PBMC", ]
    stopifnot(nrow(cur_db_ln)+nrow(cur_db_pbmc) == 
              sum(db[[col_clone_id]] %in% cur_vec_mabs_clone_id))
    
    # initialize
    cur_vec_clone_count_ln = rep(NA, length(vec_cell_type_uniq_ln))
    names(cur_vec_clone_count_ln) = vec_cell_type_uniq_ln
    cur_vec_clone_perc_ln = cur_vec_clone_count_ln
    
    cur_vec_clone_count_pbmc = rep(NA, length(vec_cell_type_uniq_pbmc))
    names(cur_vec_clone_count_pbmc) = vec_cell_type_uniq_pbmc
    cur_vec_clone_perc_pbmc = cur_vec_clone_count_pbmc
    
    if (nrow(cur_db_ln)>0) {

        cur_clone_info_ln = summarize_clone(db=cur_db_ln, col_clone=col_clone_id,
                                            col_vec=c(col_comp_tp), order_by_size="none")

        for (cur_cell_type in vec_cell_type_uniq_ln) {
            if (cur_cell_type %in% colnames(cur_clone_info_ln)) {
                cur_vec_clone_count_ln[cur_cell_type] = sum(cur_clone_info_ln[[cur_cell_type]]>0)
            }
        }
        
        cur_vec_clone_perc_ln = cur_vec_clone_count_ln/nrow(cur_clone_info_ln)*100
        
        cat("LN\n")
        print(cur_vec_clone_count_ln)
        print(cur_vec_clone_perc_ln)
        
    } else {
        cat(cur_type, ": not detected in LN\n")
        
    }
    
    if (nrow(cur_db_pbmc)>0) {
        
        cur_clone_info_pbmc = summarize_clone(db=cur_db_pbmc, col_clone=col_clone_id,
                                              col_vec=c(col_comp_tp), order_by_size="none")

        for (cur_cell_type in vec_cell_type_uniq_pbmc) {
            if (cur_cell_type %in% colnames(cur_clone_info_pbmc)) {
                cur_vec_clone_count_pbmc[cur_cell_type] = sum(cur_clone_info_pbmc[[cur_cell_type]]>0)
            }
        }
        
        cur_vec_clone_perc_pbmc = cur_vec_clone_count_pbmc/nrow(cur_clone_info_pbmc)*100
        
        cat("PBMC\n")
        print(cur_vec_clone_count_pbmc)
        print(cur_vec_clone_perc_pbmc)
        
    } else {
        cat(cur_type, ": not detected in PBMC\n")
    }
    
    # +1 for space in between LN and PBMC
    vec_x_pos = 1:(length(vec_cell_type_uniq_ln)+length(vec_cell_type_uniq_pbmc)+1)
    vec_x_label = c(vec_cell_type_uniq_ln_label, "", vec_cell_type_uniq_pbmc_label)
    vec_x_color = c(vec_cell_type_uniq_ln_color, NA, vec_cell_type_uniq_pbmc_color)
    vec_y_val = c(cur_vec_clone_perc_ln, NA, cur_vec_clone_perc_pbmc)
    
    cur_plot_fn = paste0("fig_xx_", cur_type, ".pdf")
    pdf(cur_plot_fn, width=5, height=5)
    par(mar=c(6,4.5,1,1)) # BLTR
    plot(x=vec_x_pos, y=vec_y_val,
         ylim=c(0, 100),
         type="h", lwd=10, bty="n", # no box
         xaxt="n", xlab="", ylab="% of clones", 
         col=vec_x_color, main=cur_title)
    axis(side=1, at=vec_x_pos, labels=vec_x_label, tick=F, las=2)
    dev.off()
}
```


# UMAPs

#### Prep

```{r}
# UMAP coordinates
gex_umap_zz = gzfile(fn_umap,'rt')
gex_umap = read.table(gex_umap_zz, header=T, sep="\t", stringsAsFactors=F)
close(gex_umap_zz)

# match up cells in `db` with those in `gex_umap`
idx = match(db[["cell_id"]], gex_umap[["cell_id"]])

bool_idx_non_na = !is.na(idx)

# sanity check
stopifnot(all.equal( db[["cell_id"]][bool_idx_non_na],
                     gex_umap[["cell_id"]][ idx[bool_idx_non_na] ] ))

# create a column indicating whether a cell in `gex_umap` belonged to 
# a S+ B cell clone
gex_umap[[col_s_pos_clone]] = FALSE
gex_umap[[col_s_pos_clone]][ idx[bool_idx_non_na] ] = TRUE

gex_umap[[col_donor]] = factor(gex_umap[[col_donor]], levels=DONORS)

# the ordering of the points in a ggplot2 plot depends on 
# the ordering of the rows of the input data
# later rows are plotted later

# put $s_pos_clone TRUE after FALSE
# order() not affected by factor levels
# with order(), FALSE comes before TRUE
gex_umap = gex_umap[order(gex_umap[[col_s_pos_clone]]), ]

gex_umap[[col_s_pos_clone]] = factor(gex_umap[[col_s_pos_clone]],
                                     levels=c(TRUE, FALSE))

# visualization configs

vec_gg_anno_col = c("GC"="dodgerblue2",
                    "LNPC"="forestgreen",
                    "MBC"="plum2",
                    "Naive"="darkgoldenrod2")

stopifnot(all(gex_umap[[col_anno]] %in% names(vec_gg_anno_col)))

umap_xlim = range(gex_umap[["umap_x"]]) * c(1.05, 1.05)
umap_ylim = range(gex_umap[["umap_y"]]) * c(1.05, 1.05)

g_size_s_pos = 1.25
```


#### Fig 2A (right) - S+

```{r}
g = ggplot(data=gex_umap,
           mapping=aes(x=umap_x, y=umap_y)) +
    geom_point(aes(color=s_pos_clone, size=s_pos_clone), 
               alpha=0.7, stroke=0) +
    scale_color_manual(values=c("TRUE"="red", "FALSE"="gray")) +
    scale_size_manual(values=c("TRUE"=g_size_s_pos, "FALSE"=1)) +
    xlim(umap_xlim) +
    ylim(umap_ylim) +
    theme_bw() +
    theme(axis.title.x=element_blank(),   
          axis.text.x=element_blank(),    
          axis.ticks.x=element_blank()) + 
    theme(axis.title.y=element_blank(),   
          axis.text.y=element_blank(),    
          axis.ticks.y=element_blank()) + 
    theme(legend.position="none") + 
    theme(panel.grid.major=element_blank()) + 
    theme(panel.grid.minor=element_blank())  

fn = "fig_2A_right.png"
ggsave(filename=fn, plot=g, bg="transparent", 
       device="png", width=4, height=4, units="in")
rm(g)
```


#### Fig S2E - S+, by participant

```{r}
g = ggplot(data=gex_umap,
           mapping=aes(x=umap_x, y=umap_y)) +
    geom_point(aes(color=.data[[col_s_pos_clone]], size=.data[[col_s_pos_clone]]), 
               alpha=0.7, stroke=0) +
    scale_color_manual(values=c("TRUE"="red", "FALSE"="gray")) +
    scale_size_manual(values=c("TRUE"=1.2, "FALSE"=0.8)) +
    facet_wrap(facets=vars(.data[[col_donor]]), nrow=2) +
    theme_bw() +
    theme(strip.background=element_rect(fill="transparent"),
          strip.text=element_text(size=13)) +
    theme(axis.title.x=element_blank(),   
          axis.text.x=element_blank(),    
          axis.ticks.x=element_blank()) + 
    theme(axis.title.y=element_blank(),   
          axis.text.y=element_blank(),    
          axis.ticks.y=element_blank()) + 
    theme(legend.position="none") + 
    theme(panel.grid.major=element_blank()) 

fn = "fig_S2E.png"
ggsave(filename=fn, plot=g, bg="transparent", 
       device="png", width=9, height=6, units="in")
rm(g)
```


