---
title: "Matz et al., *Nature Immunology*, 2026 (repertoire)"
output:
  html_document:
    df_print: paged
---

Author: Julian Q. Zhou

Docker container: `julianqz/wu_pub:r_4.1.0`

Visualizations are outputted as pdfs or pngs.

# Load packages & helper functions

```{r}
# modify the next line to set working directory accordingly
# setwd("[insert_your_working_directory_here]")

# summarize_clone
source("./c2b2/bcr_clone_info.R")

# circos_clonal_overlap
source("./c2b2/bcr_clonal_overlap_circos_v2.R")

# barplot_2
source("./c2b2/vis_barplot.R")
```

```{r}
library(tibble)
library(ggplot2)

library(ape)
library(ggtree)
```

```{r}
versions::installed.versions(c("ggplot2", "circlize", "ape", "ggtree"))
```


# Setup

```{r}
### filenames
fn_db_hc = "WU397_matz_et_al_nat_imm_2026_bcr_heavy.tsv.gz"
fn_db_lc = "WU397_matz_et_al_nat_imm_2026_bcr_light.tsv.gz"
fn_umap = "WU397_matz_et_al_nat_imm_2026_gex_b_cell_umap.tsv.gz"

### column names

col_comp = "compartment"
col_tp = "timepoint_2" # timepoint in days
col_tp_w = "timepoint_2_wk" # timepoint in weeks

# [compartment]_[timepoint]
col_comp_tp = "compartment_timepoint_2"

# B cell clonal lineage ID
col_clone_id = "clone_id"

# ID of expressed mAb
col_exp_id = "expressed_id"

# "pos" if an expressed mAb was tested positive for HA binding via ELISA
col_ha_binary = "ha_binding_binary" 

# TRUE if clone contains HA+
col_ha_pos_clone = "ha_pos_clone"

# transcriptomics-based cluster annotation
col_anno = "anno_leiden_0.95"

# replacement and silent mutation frequency
col_mf = "nuc_RS_freq_1_312"

col_donor = "donor"
col_sample = "sample"
col_seq_id = "sequence_id"
col_cell_id = "cell_id"
col_iso = "isotype"
col_tissue = "tissue"

### participants
DONORS = paste0("397-", c("06","17","28","29","05","09", "22"))
```


# Load BCR data

```{r}
# db_all contains heavy chains from all clones regardless of S specificity
db_all_hc_zz = gzfile(fn_db_hc,'rt')
db_all = read.table(db_all_hc_zz, header=T, sep="\t", stringsAsFactors=F)
close(db_all_hc_zz)

# clones containing HA-positive expressed mAbs
vec_clones_ha_pos = unique(db_all[[col_clone_id]][db_all[[col_ha_binary]]=="pos"])

# Subset to HA+ clones
db = db_all[db_all[[col_clone_id]] %in% vec_clones_ha_pos, ]
rm(db_all)

# light chain
db_all_lc_zz = gzfile(fn_db_lc,'rt')
db_all_lc = read.table(db_all_lc_zz, header=T, sep="\t", stringsAsFactors=F)
close(db_all_lc_zz)
```

```{r}
# Recalculate SHM frequency to avoid numerical problems affecting $nuc_RS_freq_1_312
# that arise from writing and reading tsv's

# E.g. a "0.013108614" could become "0.013108615" after being exported and read back in

# Doing so ensures that the exact p-values reported in the paper can be reproduced
db[[col_mf]] = db[["nuc_RS_1_312"]] / db[["nuc_denom_1_312"]]
db_all_lc[[col_mf]] = db_all_lc[["nuc_RS_1_312"]] / db_all_lc[["nuc_denom_1_312"]]
```


# Clone-level summary

First, we create a `data.frame` in which each row represents a clone, and columns
tabulate the number of sequences in a compartment and in a compartment-timepoint
combination.

```{r}
clone_info = summarize_clone(db=db, col_clone=col_clone_id, 
                             col_vec=c(col_comp, col_comp_tp),
                             order_by_size="none")

# Extract participant ID from clone ID, which has the format [donor]@[clone]
clone_info[[col_donor]] = sapply(clone_info[[col_clone_id]],
                                 function(s){strsplit(s, "@")[[1]][1]})
```


# Circos plots for clonal overlap

`circos_clonal_overlap_v2` produces a circos plot, and prints out (i) the number and
percentage of clones with clonal overlap, and (ii) the number of sequences.

Starting with the arc at the 12 o'clock position and going clockwise, the identities of 
the arcs correspond to the rows in the printed table from top to bottom.

### How to read the circos plots

* An arc represents BCR sequences from a B cell compartment or 
compartment-timepoint combination
* The length of an arc corresponds to the number of BCR sequences 
represented by the arc
* Chords connect clonally related sequences
* The width of a chord corresponds to the size of the clone represented 
by the chord
* Chords are coloured based on whether there's "overlap"
* Overlap can be defined as overlap between compartments, between 
compartment-timepoint combinations, etc.

#### Fig 3a - clonal overlap between GC B cell compartment and w1 PB

The percentages reported in Fig 3a were extracted from `fig_3a_info.txt` produced below.

```{r}
### prep

vec_sectors_comp_tp = c("PB_d8", "GC")
vec_sectors_type_comp_tp = c("PB","GC")
vec_sectors_color_comp_tp = c("PB"="firebrick2", "GC"="dodgerblue2")

vec_color_overlap = c("GC@PB"="purple")
color_rest = "gray40"

vec_gaps_comp_tp = rep(4, length(vec_sectors_comp_tp))

col_sector_comp_tp = "tmp_circos_col"


fn_sink = "fig_3a_info.txt"
sink(fn_sink)

for (d in DONORS) {
    cat("\n", d, "\n")
    # subset to current donor
    d_db = db[db[[col_donor]]==d, ]
    d_clone_info = clone_info[clone_info[[col_clone_id]] %in% d_db[[col_clone_id]], ]
    
    # tmp column
    # this is a workaround for GCB having timepoints combined and PB having timepoint specified
    # make a copy of the $compartment column
    # GC remains GC, without timepoint specified
    d_db[[col_sector_comp_tp]] = d_db[[col_comp]]
    # change value for PB to PB_d8 (i.e. specifying timepoint)
    d_db[[col_sector_comp_tp]][which(d_db[[col_comp_tp]]=="PB_d8")] = "PB_d8"
    
    d_fn = paste0("fig_3a_", d, ".pdf")
    pdf(d_fn, width=5, height=5)
    
    circos_clonal_overlap_v2(vec_sectors=vec_sectors_comp_tp, 
                             vec_sectors_type=vec_sectors_type_comp_tp, 
                             vec_sectors_color=vec_sectors_color_comp_tp, 
                             vec_gaps=vec_gaps_comp_tp,  
                             col_sector=col_sector_comp_tp, 
                             col_clone_id=col_clone_id, 
                             df_seq_data=d_db, 
                             df_clone_info=d_clone_info, 
                             color_overlap=vec_color_overlap,
                             color_rest=color_rest,
                             color_mono=FALSE)
    
    dev.off()
}

sink()
```


# Clone-wise SHM frequency comparisons between compartment-timepoint combinations

*The code-generated plots from below were prototypes based on which the published 
versions were generated using GraphPad Prism v10.*

##### How to read the clone-wise SHM frequency plots

* Each line corresponds to a B cell clone
* Each end of the line corresponds to the SHM frequency of a clonal 
member in a given compartment or compartment-timepoint combination that
was expressed as a mAb and verified to be HA-positive via ELISA


#### Fig 3b - wk 1 PB vs. w26 GC B cells

```{r}
# clones spanning wk 1 PB and wk 26 GC B cell compartments
vec_cl = clone_info[[col_clone_id]][ clone_info[["PB_d8"]]>0 & clone_info[["GC_d180"]]>0  ]

# placeholder
vec_cols_df_mab_shm = c(col_clone_id, 
                        "mab_id_PB_d8", "mab_id_GC_d180",
                        "shm_hc_PB_d8", "shm_lc_PB_d8", 
                        "shm_hc_GC_d180", "shm_lc_GC_d180")
df_mab_shm = data.frame(matrix(NA, nrow=length(vec_cl), ncol=length(vec_cols_df_mab_shm)))
colnames(df_mab_shm) = vec_cols_df_mab_shm

# fill in values
df_mab_shm[[col_clone_id]] = vec_cl

for (i_cl in 1:nrow(df_mab_shm)) {
    cur_cl = df_mab_shm[[col_clone_id]][i_cl]
    
    # locate heavy chain; wrt db
    idx_hc_pb_d8 = which(db[[col_clone_id]]==cur_cl &
                         db[[col_exp_id]]!="" & # expressed mAb
                         db[[col_comp_tp]]=="PB_d8")
    stopifnot(length(idx_hc_pb_d8)==1)
    
    idx_hc_gc_d180 = which(db[[col_clone_id]]==cur_cl &
                         db[[col_exp_id]]!="" & # expressed mAb
                         db[[col_comp_tp]]=="GC_d180")
    stopifnot(length(idx_hc_gc_d180)==1)
    
    # locate corresponding light chain; wrt db_all_lc
    idx_lc_pb_d8 = which(db_all_lc[[col_cell_id]]==db[[col_cell_id]][idx_hc_pb_d8])
    stopifnot(length(idx_lc_pb_d8)==1)
    
    idx_lc_gc_d180 = which(db_all_lc[[col_cell_id]]==db[[col_cell_id]][idx_hc_gc_d180])
    stopifnot(length(idx_lc_gc_d180)==1)
    
    # mAb ID; derive from HC expressed ID by removing ".HC" at the end
    df_mab_shm[["mab_id_PB_d8"]][i_cl] = strsplit(db[[col_exp_id]][idx_hc_pb_d8], ".HC")[[1]][1]
    df_mab_shm[["mab_id_GC_d180"]][i_cl] = strsplit(db[[col_exp_id]][idx_hc_gc_d180], ".HC")[[1]][1]
    # heavy chain SHM frequency
    df_mab_shm[["shm_hc_PB_d8"]][i_cl] = db[[col_mf]][idx_hc_pb_d8]
    df_mab_shm[["shm_hc_GC_d180"]][i_cl] = db[[col_mf]][idx_hc_gc_d180]
    # light chain SHM frequency
    df_mab_shm[["shm_lc_PB_d8"]][i_cl] = db_all_lc[[col_mf]][idx_lc_pb_d8]
    df_mab_shm[["shm_lc_GC_d180"]][i_cl] = db_all_lc[[col_mf]][idx_lc_gc_d180]
}

# expect no NA left
stopifnot(!any(is.na(df_mab_shm)))

# vis, heavy chain

ymax_common = max(df_mab_shm[, grepl(pattern="shm_", x=colnames(df_mab_shm))])*1.05

pdf("fig_3b.pdf", width=10, height=5)
par(mfrow=c(1,2), mar=c(5,5,1,1))

plot(x=0, y=0, xlim=c(0,3), ylim=c(0, ymax_common), xaxt="n", col=NA,
     xlab="", ylab="IGHV nucleotide mutation frequency")
axis(side=1, at=c(1,2), labels=c("w1 PB", "w26 GC"))
for (i_cl in 1:nrow(df_mab_shm)) {
    points(x=c(1,2), y=c(df_mab_shm[["shm_hc_PB_d8"]][i_cl],
                         df_mab_shm[["shm_hc_GC_d180"]][i_cl]),
           col=c("firebrick2", "dodgerblue2"))
    lines(x=c(1,2), y=c(df_mab_shm[["shm_hc_PB_d8"]][i_cl],
                         df_mab_shm[["shm_hc_GC_d180"]][i_cl]),
           col=c("plum2"))
}

# vis, light chain
plot(x=0, y=0, xlim=c(0,3), ylim=c(0, ymax_common), xaxt="n", col=NA,
     xlab="", ylab="IGK/LV nucleotide mutation frequency")
axis(side=1, at=c(1,2), labels=c("w1 PB", "w26 GC"))

for (i_cl in 1:nrow(df_mab_shm)) {
    points(x=c(1,2), y=c(df_mab_shm[["shm_lc_PB_d8"]][i_cl],
                         df_mab_shm[["shm_lc_GC_d180"]][i_cl]),
           col=c("firebrick2", "dodgerblue2"))
    lines(x=c(1,2), y=c(df_mab_shm[["shm_lc_PB_d8"]][i_cl],
                         df_mab_shm[["shm_lc_GC_d180"]][i_cl]),
           col=c("plum2"))
}

dev.off()

# the reported p-values were from tests carried out in Prism
# below is code for running the tests in R
wilcox.test(x=df_mab_shm[["shm_hc_PB_d8"]], y=df_mab_shm[["shm_hc_GC_d180"]], paired=T)
wilcox.test(x=df_mab_shm[["shm_lc_PB_d8"]], y=df_mab_shm[["shm_lc_GC_d180"]], paired=T)


```

# Isotype

#### Fig 3e

```{r}
# vis config
uniq_iso = paste0("IGH", c("G","A","M","D","E"))
# order should match uniq_iso in order for barplot_2 to work properly
uniq_iso_colors = c("IGHG"="hotpink",
                    "IGHA"="forestgreen", 
                    "IGHM"="dodgerblue2", 
                    "IGHD"="darkgoldenrod2", 
                    "IGHE"="purple")

# subset data for vis
vec_cl_gc = clone_info[[col_clone_id]][ clone_info[["GC"]]>0 ]
clone_info_iso = clone_info[clone_info[[col_clone_id]] %in% vec_cl_gc, ]
db_iso = db[db[[col_clone_id]] %in% vec_cl_gc, ]
    
# out of all cells vs. GC B cells 
for (cur_mode in c("all_cells", "gcb_only")) {
    
    if (cur_mode=="gcb_only") {
        # subset to GC B cells only
        db_iso = db_iso[which(db_iso[[col_comp]]=="GC"), ]
        
        # only need to do this if subsetting to gcb_only
        clone_info_iso = summarize_clone(db=db_iso, col_clone=col_clone_id,
                                         col_vec=c(col_iso), order_by_size="none")
        clone_info_iso[[col_donor]] = sapply(clone_info_iso[[col_clone_id]],
                                             function(x){strsplit(x, "@")[[1]][1]}, USE.NAMES=F)
    }
    
    # calculate isotype distribution
    iso_cell_by_donor = t(sapply(DONORS,
                                 function(d){
                                     return(sapply(uniq_iso, 
                                                   function(iso){
                                                       sum(db_iso[[col_iso]]==iso & db_iso[[col_donor]]==d)
                                                   }))
                                 }))
    print(iso_cell_by_donor)
    
    # convert from counts to percentages
    iso_cell_by_donor_perc = t(apply(iso_cell_by_donor, 1, function(x){ x/sum(x)*100 }))
    # expect all 100 (or NaN)
    rowSums(iso_cell_by_donor_perc)    
    print(round(iso_cell_by_donor_perc, 1))
    
    # vis
    fn = paste0("fig_3e_", cur_mode, ".pdf")
    pdf(fn, width=6, height=5)
    par(mar=c(5,4.5,0,0)) # BLTR
    barplot_2(tab=iso_cell_by_donor_perc, 
              vec_ylim=c(0,105), col_bar=uniq_iso_colors, ylab="%")
    dev.off()
}



```

# UMAPs

#### Prep

```{r}
# UMAP coordinates
gex_umap_zz = gzfile(paste0("./assets/", fn_umap),'rt')
gex_umap = read.table(gex_umap_zz, header=T, sep="\t", stringsAsFactors=F)
close(gex_umap_zz)

# simplify timepoint: d180 and d181 both treated as d180
gex_umap[[col_tp]] = gex_umap[["timepoint"]]
gex_umap[[col_tp]][gex_umap[[col_tp]]=="d181"] = "d180"

# factorize timepoint
gex_umap[[col_tp]] = factor(gex_umap[[col_tp]], levels=paste0("d", c(0,8,15,57,121,180)))

# match up cells in `db` with those in `gex_umap`
idx = match(db[[col_cell_id]], gex_umap[[col_cell_id]])

bool_idx_non_na = !is.na(idx)

# sanity check
stopifnot(all.equal( db[[col_cell_id]][bool_idx_non_na],
                     gex_umap[[col_cell_id]][ idx[bool_idx_non_na] ] ))

# create a column indicating whether a cell in `gex_umap` belonged to 
# a HA+ B cell clone
gex_umap[[col_ha_pos_clone]] = FALSE
gex_umap[[col_ha_pos_clone]][ idx[bool_idx_non_na] ] = TRUE

gex_umap[[col_donor]] = factor(gex_umap[[col_donor]], levels=DONORS)

# the ordering of the points in a ggplot2 plot depends on 
# the ordering of the rows of the input data
# later rows are plotted later

# put $ha_pos_clone TRUE after FALSE
# order() not affected by factor levels
# with order(), FALSE comes before TRUE
gex_umap = gex_umap[order(gex_umap[[col_ha_pos_clone]]), ]

gex_umap[[col_ha_pos_clone]] = factor(gex_umap[[col_ha_pos_clone]],
                                     levels=c(TRUE, FALSE))

# visualization configs
vec_gg_anno_col = c("ABC"="mediumspringgreen",
                    "GC"="dodgerblue2",
                    "LNPC"="forestgreen",
                    "MBC"="plum2",
                    "Naive"="darkgoldenrod2",
                    "PB"="firebrick2")

stopifnot(all(gex_umap[[col_anno]] %in% names(vec_gg_anno_col)))
```

#### Fig 2c

```{r}
g = ggplot(data=gex_umap,
           mapping=aes(x=umap_x, y=umap_y)) +
    geom_point(aes_string(color=col_anno), 
               alpha=0.7, stroke=0, size=1) +
    scale_color_manual(values=vec_gg_anno_col) +
    facet_grid(rows=vars(.data[[col_donor]]), cols=vars(.data[[col_tp]])) +
    theme_bw() +
    theme(strip.background=element_rect(fill="transparent"),
          strip.text=element_text(size=13)) +
    theme(axis.title.x=element_blank(),   # no x-axis title
          axis.text.x=element_blank(),    # no x-axis tick label
          axis.ticks.x=element_blank()) + # no x-axis ticks 
    theme(axis.title.y=element_blank(),   # no y-axis title
          axis.text.y=element_blank(),    # no y-axis tick label
          axis.ticks.y=element_blank()) + # no y-axis ticks 
    theme(legend.position="none") + # no legend
    theme(panel.grid.major=element_blank()) #+ # remove major grid

fn = "fig_2c.png"
ggsave(filename=fn, plot=g, bg="transparent", 
       device="png", 
       width=15, height=12, units="in")  
```

#### Fig 2d & Extended Data Fig 3e

```{r}
g = ggplot(data=gex_umap,
           mapping=aes(x=umap_x, y=umap_y)) +
    geom_point(aes(color=.data[[col_ha_pos_clone]], size=.data[[col_ha_pos_clone]]), 
               alpha=0.7, stroke=0) +
    scale_color_manual(values=c("TRUE"="red", "FALSE"="gray")) +
    scale_size_manual(values=c("TRUE"=1, "FALSE"=0.8)) +
    facet_grid(rows=vars(.data[[col_donor]]), cols=vars(.data[[col_tp]])) +
    theme_bw() +
    theme(strip.background=element_rect(fill="transparent"),
          strip.text=element_text(size=13)) +
    theme(axis.title.x=element_blank(),   # no x-axis title
          axis.text.x=element_blank(),    # no x-axis tick label
          axis.ticks.x=element_blank()) + # no x-axis ticks 
    theme(axis.title.y=element_blank(),   # no y-axis title
          axis.text.y=element_blank(),    # no y-axis tick label
          axis.ticks.y=element_blank()) + # no y-axis ticks 
    theme(legend.position="none") + # no legend
    theme(panel.grid.major=element_blank()) # remove major grid

fn = "fig_2d_ed_fig_3e.png"
ggsave(filename=fn, plot=g, bg="transparent", 
       device="png",
       width=15, height=12, units="in")
```

# Phylogenetic trees

#### Extended Data Fig 4c

```{r}
# files containing tree topology 
# each file corresponds to a HA+ B cell clone
vec_tree_files = list.files(path="./assets/igphyml_output", pattern="_tree_HLP.txt")
vec_tree_clones = sapply(vec_tree_files, function(s){ strsplit(s, ".fasta")[[1]][1] })

# sanity check that all the clones are in `db`
stopifnot(all(vec_tree_clones %in% db[[col_clone_id]]))

# timepoint conversion: days to weeks
vec_tp_convert = c("d0"="w0", "d8"="w1", "d15"="w2", "d57"="w8", "d121"="w17", "d180"="w26")
db[[col_tp_w]] = vec_tp_convert[db[[col_tp]]]

# loop through the clones
for (i_cl in 1:length(vec_tree_clones)) {
    
    cur_cl = vec_tree_clones[i_cl]
    cur_cl_split = strsplit(cur_cl, "@")[[1]]
    # donor ID
    cur_d = cur_cl_split[1]
    
    cur_fn_tree = vec_tree_files[i_cl]
    cur_fn_plot = paste0("ed_fig_4c_", i_cl, "_", cur_cl, ".pdf")
    
    ### read in tree
    
    cur_newick = read.tree(file=paste0("./assets/igphyml_output/", cur_fn_tree))
    # phylo
    class(cur_newick)
    
    # number of tips
    cur_n_tips = length(cur_newick[["tip.label"]])
    
    
    ### metadata
    
    cur_tibble = as_tibble(cur_newick)
    # tbl_tree
    class(cur_tibble)
    
    # extract sequence_id to use as anchor to obtain metadata from `db`
    cur_tip_ids = sapply(cur_tibble[["label"]], function(s){
        if (!is.na(s)) {
            # if _GERM_
            if (grepl(pattern="_GERM_", x=s)) {
                return("Germline")
            } else {
                # from: 397-05_s26@CCGTACTTCCTATTCA-1_contig_2_397-05_s26_LN_IGHA_GC_d15_NA
                    # to: 397-05_s26@CCGTACTTCCTATTCA-1_contig_2
                    ss = strsplit(x=s, split=paste0("_", cur_d, "_"))[[1]][1]
                    return(ss)
            }
        } else {
            # NA remains NA
            return(NA)
        }
    })
    
    # match against db
    cur_idx = match(cur_tip_ids, db[[col_seq_id]])
    # sanity check
    # NAs in cur_idx correspond to NAs or _GERM_ in $label
    # +1 for _GERM_
    stopifnot(sum(is.na(cur_idx)) == sum(is.na(cur_tibble[["label"]]))+1)
    
    bool_cur_idx_no_na = !is.na(cur_idx)
    # sanity check
    stopifnot(all.equal(cur_tip_ids[bool_cur_idx_no_na], 
                        db[[col_seq_id]][cur_idx[bool_cur_idx_no_na]],
                        check.attributes=F))
    
    # meta columns
    cur_meta_cols = c(col_donor, col_sample, col_tissue, col_iso, 
                      col_comp, col_tp, col_tp_w, col_ha_binary)
    stopifnot(all(cur_meta_cols %in% colnames(db)))
    
    cur_meta = db[cur_idx[bool_cur_idx_no_na], cur_meta_cols]
    
    for (cur_col in cur_meta_cols) {
        cur_tibble[[cur_col]] = NA
        cur_tibble[[cur_col]][bool_cur_idx_no_na] = cur_meta[[cur_col]]
    }
    
    # aesthetic mapping
    # colour: compartment (PB, GC, LNPC)
    # shape: isotype 
    
    vec_map_fill = c("PB"="firebrick2", "GC"="dodgerblue2", "LNPC"="forestgreen",
                         "MBC_blood"="plum2", "MBC_LN"="purple3", "ABC"="seagreen2")
        
    # 23 diamond
    # 21 circle
    # 22 square
    vec_map_shape = c("IGHA"=23, "IGHG"=21, "IGHM"=22)
    
    stopifnot(all(names(vec_map_fill) %in% db[[col_comp]]))
        
    stopifnot(all(names(vec_map_shape) %in% db[[col_iso]]))
    
    # factorize for ggplot2
    cur_tibble[[col_comp]] = factor(cur_tibble[[col_comp]], levels=names(vec_map_fill))
    cur_tibble[[col_iso]] = factor(cur_tibble[[col_iso]], levels=names(vec_map_shape)) 
    
    head(cur_tibble)
    
    # back to phylo
    cur_phylo = as.phylo(cur_tibble)
    
    
    ### visualize

    # There's a row with all NA's
    # This corresponds to the germline
    # Its presence triggers a warning:
    # "Removed 1 rows containing missing values (geom_point_g_gtree)"
    # This warning can be safely ignored
    
    # tip symbol and label sizes by trial and error
    size_tippoint = c(7,4) 
    size_text_lab = c(3,2)
    height_pdf=5
    
    p_base = ggtree(cur_phylo)
    
    p_comp_tp = p_base %<+% cur_tibble +
        # `stroke` controls width of outline of shape 
        # `size` controls area of inside of shape
        geom_tippoint(aes(shape=isotype, 
                          fill=compartment),
                      size=size_tippoint[i_cl], stroke=0, col="transparent") +
        # `na.translate=F` hides the NA level in legend
        # `name` controls title of respective legend
        scale_shape_manual(
            values=vec_map_shape[names(vec_map_shape) %in% cur_tibble[[col_iso]]], 
            na.translate=F, name="Isotype") +
        scale_fill_manual(
            values=vec_map_fill[names(vec_map_fill) %in% cur_tibble[[col_comp]]], 
            na.translate=F, name="Compartment") +
        # guides(size=F) + # hide legend for `size`
        # override aes as default doesn't appear to be working
        # order specified position of each of multiple legends (`order` should be outside `override.aes`)
        guides(fill=guide_legend(
            override.aes=list(size=5, 
                              color=vec_map_fill[names(vec_map_fill) %in% cur_tibble[[col_comp]]]),
            order=2),
            shape=guide_legend(
                override.aes=list(size=4, stroke=1, color="black", fill="transparent",
                                  shape=vec_map_shape[names(vec_map_shape) %in% cur_tibble[[col_iso]]]),
                order=1)
        ) +
        geom_tiplab(aes(label=timepoint_2_wk), hjust=-0.8, size=size_text_lab[i_cl]) + 
        geom_treescale(x=0, y=cur_n_tips*0.9) 
    
    
    ggsave(filename=cur_fn_plot, plot=p_comp_tp, device="pdf", width=5, height=height_pdf)
    
    rm(p_base, p_comp_tp, cur_tibble, cur_meta, cur_phylo)
}
```

